---
title: "Get data for fitting a robust model"
subtitle: "Filter out samples with missing aha8_total scores, or that have reached dimentia, and add cell type proportion PCs"
author: 
  - David lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)
  library(recountmethylation)
  library(SummarizedExperiment)
})

base_dir <- "~/TBL Dropbox/David Lukacsovich"
analysis_dir <- file.path(base_dir, "AD_CR_NOMAS")
data_dir <- file.path(analysis_dir, "data/DNAm/processed")
result_dir <- file.path(analysis_dir, "analysis-results")
ancestry_dir <- file.path(result_dir, "ancestry")
glint_dir <- file.path(analysis_dir, "code", "GLINT_1.0.4")
```

## Read in Data

```{r get_data}
dnam_112_data <- readRDS(file.path(data_dir, "EPIC112_Batched_PCA.RDS"))
dnam_208_data <- readRDS(file.path(data_dir, "EPIC208_Batched_PCA.RDS"))
```

# Get Parameters

## Filter Data

```{r filt}
message(
  "There are ",
  ncol(dnam_112_data), " EPIC112 samples, and ",
  ncol(dnam_208_data), " EPIC208 samples."
)

dnam_112_data <- dnam_112_data[,!is.na(dnam_112_data$aha8_total)]
dnam_208_data <- dnam_208_data[,!is.na(dnam_208_data$aha8_total)]
message(
  "After filtering samples with missing aha8_total values, there are ",
  ncol(dnam_112_data), " EPIC112 samples, and ",
  ncol(dnam_208_data), " EPIC208 samples."
)

dnam_112_data <- dnam_112_data[,!is.na(dnam_112_data$apoe4)]
dnam_208_data <- dnam_208_data[,!is.na(dnam_208_data$apoe4)]
message(
  "After filtering samples with missing apoe4 values, there are ",
  ncol(dnam_112_data), " EPIC112 samples, and ",
  ncol(dnam_208_data), " EPIC208 samples."
)
```

## Calculate aha8 and time difference parameters

```{r}
dnam_112_data$aha8_zscore <- scale(
  dnam_112_data$aha8_total, center = TRUE, scale = TRUE
)
dnam_208_data$aha8_zscore <- scale(
  dnam_208_data$aha8_total, center = TRUE, scale = TRUE
)

dnam_112_data$diff_time <- dnam_112_data$age_at_visit - dnam_112_data$np1age
dnam_208_data$diff_time <- dnam_208_data$age_at_visit - dnam_208_data$np1age
```

# Get Celltype PCA

## Define Functions

```{r func_pca}
get_dnam_pca <- function(dnam_data) {
  pheno_df <- dnam_data %>%
    SummarizedExperiment::colData() %>%
    as.data.frame() %>%
    dplyr::select(
      "B", "NK", "CD4T", "CD8T", "Mono", "Neutro", "Eosino"
    )
  
  pc_vals <- stats::prcomp(pheno_df, center = TRUE, scale = TRUE)
  
  pc_vals
}

get_pca_statistics <- function(pc_vals) {
  list(
    pc1_mean = mean(pc_vals$x[,1]),
    pc2_mean = mean(pc_vals$x[,2]),
    pc3_mean = mean(pc_vals$x[,3]),
    pc1_sd = stats::sd(pc_vals$x[,1]),
    pc2_sd = stats::sd(pc_vals$x[,2]),
    pc3_sd = stats::sd(pc_vals$x[,3])
  )
}

plot_dnam_pca <- function(pc_vals, title) {
  plot_df <- data.frame(pc_vals$x[,1:3])
  colnames(plot_df) <- paste0("PC", 1:3)
  
  percentVar <- round(100 * pc_vals$sdev ** 2 / sum(pc_vals$sdev ** 2), 1)
  pca_params <- get_pca_statistics(pc_vals)
  
  outlier_cutoff <- 3
  
  p1 <- ggplot2::ggplot(
    data = plot_df, mapping = ggplot2::aes(x = PC1, y = PC2)
  ) +
    ggplot2::geom_point(size = 1) +
    ggplot2::theme_bw() +
    ggplot2::xlab(paste0("PC1: ", percentVar[[1]], "% variance")) +
    ggplot2::ylab(paste0("PC2: ", percentVar[[2]], "% variance")) +
    ggplot2::geom_hline(
      yintercept = pca_params$pc2_mean + outlier_cutoff * pca_params$pc2_sd,
      linetype = "dashed"
    ) +
    ggplot2::geom_hline(
      yintercept = pca_params$pc2_mean - outlier_cutoff * pca_params$pc2_sd,
      linetype = "dashed"
    ) +
    ggplot2::geom_vline(
      xintercept = pca_params$pc1_mean + outlier_cutoff * pca_params$pc1_sd,
      linetype = "dashed"
    ) +
    ggplot2::geom_vline(
      xintercept = pca_params$pc1_mean - outlier_cutoff * pca_params$pc1_sd,
      linetype = "dashed"
    ) +
    ggplot2::ggtitle(title)
  
  p2 <- ggplot2::ggplot(
    data = plot_df, mapping = ggplot2::aes(x = PC1, y = PC3)
  ) +
    ggplot2::geom_point(size = 1) +
    ggplot2::theme_bw() +
    ggplot2::xlab(paste0("PC1: ", percentVar[[1]], "% variance")) +
    ggplot2::ylab(paste0("PC3: ", percentVar[[3]], "% variance")) +
    ggplot2::geom_hline(
      yintercept = pca_params$pc3_mean + outlier_cutoff * pca_params$pc3_sd,
      linetype = "dashed"
    ) +
    ggplot2::geom_hline(
      yintercept = pca_params$pc3_mean - outlier_cutoff * pca_params$pc3_sd,
      linetype = "dashed"
    ) +
    ggplot2::geom_vline(
      xintercept = pca_params$pc1_mean + outlier_cutoff * pca_params$pc1_sd,
      linetype = "dashed"
    ) +
    ggplot2::geom_vline(
      xintercept = pca_params$pc1_mean - outlier_cutoff * pca_params$pc1_sd,
      linetype = "dashed"
    ) +
    ggplot2::ggtitle(title)
  
  list(p1 = p1, p2 = p2)
}

run_pca_analysis <- function(dnam_data, title) {
  pc_vals <- get_dnam_pca(dnam_data)
  percentVar <- round(100 * pc_vals$sdev ** 2 / sum(pc_vals$sdev ** 2), 1)
  
  dnam_data$CPC_1 <- pc_vals$x[,1]
  dnam_data$CPC_2 <- pc_vals$x[,2]
  dnam_data$CPC_3 <- pc_vals$x[,3]
  
  plot_ls <- plot_dnam_pca(pc_vals, title)
  
  list(dnam_data = dnam_data, p1 = plot_ls$p1, p2 = plot_ls$p2, percentVar = percentVar)
}
```

## Get PCA

```{r pca_run}
res_112 <- run_pca_analysis(dnam_112_data, "EPIC112")
dnam_112_data <- res_112$dnam_data

res_208 <- run_pca_analysis(dnam_208_data, "EPIC208")
dnam_208_data <- res_208$dnam_data

message(
  "The variance explained by the first 3 PCs is ",
  sum(res_112$percentVar[1:3]), "% in EPIC112 and ",
  sum(res_208$percentVar[1:3]), "% in EPIC208."
)
```

## Plot PCA

```{r plot, fig.height = 6, fig.width = 12}
gridExtra::grid.arrange(res_112$p1, res_112$p2, ncol = 2)
gridExtra::grid.arrange(res_208$p1, res_208$p2, ncol = 2)
```

# Save Filtered Data

```{r save}
saveRDS(dnam_112_data, file.path(data_dir, "EPIC112_aha8.RDS"))
saveRDS(dnam_208_data, file.path(data_dir, "EPIC208_aha8.RDS"))
```

# Get Ancestry Predictions

## Get target cpgs

```{r get_cpgs}
fname_path <- file.path(
  system.file(package = "recountmethylation"),
  "extdata", "glint_files", "glint_epistructure_explanatory-cpgs.rda"
)
cpgs_ls <- get(load(fname_path))
```

## Write EPIC112 Files

```{r write_112}
pheno_112_df <- dnam_112_data %>%
  SummarizedExperiment::colData() %>%
  as.data.frame() %>%
  dplyr::select(
    age = "age_at_visit",
    sex = "Gender"
  ) %>%
  dplyr::mutate(
    age = as.numeric(.data$age),
    sex = ifelse(.data$sex == "M", 1, 0)
  )

beta_112 <- SummarizedExperiment::assays(dnam_112_data)$dnam
beta_112 <- beta_112[row.names(beta_112) %in% cpgs_ls,]

write.table(
  pheno_112_df,
  file = file.path(ancestry_dir, "EPIC112_covariate.txt"),
  sep = "\t",
  row.names = TRUE,
  col.names = TRUE,
  append = FALSE,
  quote = FALSE
)

write.table(
  beta_112,
  file = file.path(ancestry_dir, "EPIC112_data.txt"),
  sep = "\t",
  row.names = TRUE,
  col.names = TRUE,
  append = FALSE,
  quote = FALSE
)
```

## Write EPIC208 Files

```{r write_208}
pheno_208_df <- dnam_208_data %>%
  SummarizedExperiment::colData() %>%
  as.data.frame() %>%
  dplyr::select(
    age = "age_at_visit",
    sex = "Gender"
  ) %>%
  dplyr::mutate(
    age = as.numeric(.data$age),
    sex = ifelse(.data$sex == "M", 1, 0)
  )

beta_208 <- SummarizedExperiment::assays(dnam_208_data)$dnam
beta_208 <- beta_208[row.names(beta_208) %in% cpgs_ls,]

write.table(
  pheno_208_df,
  file = file.path(ancestry_dir, "EPIC208_covariate.txt"),
  sep = "\t",
  row.names = TRUE,
  col.names = TRUE,
  append = FALSE,
  quote = FALSE
)

write.table(
  beta_208,
  file = file.path(ancestry_dir, "EPIC208_data.txt"),
  sep = "\t",
  row.names = TRUE,
  col.names = TRUE,
  append = FALSE,
  quote = FALSE
)
```

## Write Python Commands

Writes a file called `Analysis_Commands.txt`. These commands can be run, in order, from the command terminal, to create the ancestry based pca.

```{r write_commands}
adjust_path <- function(file_path) {
  gsub(" ","\\ ", file_path,fixed = TRUE)
}

glint.pypath <- file.path(glint_dir, "glint.py")
command_112 <- paste0(
  c("python", adjust_path(glint.pypath),
    "--datafile", adjust_path(file.path(ancestry_dir, "EPIC112_data.txt")), 
    "--covarfile", adjust_path(file.path(ancestry_dir, "EPIC112_covariate.txt")), 
    "--covar", "age sex",
    "--savepcs 10",
    "--epi", "--out", adjust_path(file.path(ancestry_dir, "EPIC112_result"))
  ), 
  collapse = " "
)

command_208 <- paste0(
  c("python", adjust_path(glint.pypath),
    "--datafile", adjust_path(file.path(ancestry_dir, "EPIC208_data.txt")), 
    "--covarfile", adjust_path(file.path(ancestry_dir, "EPIC208_covariate.txt")), 
    "--covar", "age sex",
    "--savepcs 10",
    "--epi", "--out", adjust_path(file.path(ancestry_dir, "EPIC208_result"))
  ), 
  collapse = " "
)

commands <- c(
  "conda activate glint_env",
  command_112,
  command_208,
  "conda deactivate"
)

fileConn <- file(file.path(ancestry_dir, "Analysis_Commands.txt"))
writeLines(commands, fileConn)
close(fileConn)
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>
